---
title: "beta_binomial_example"
author: "Isaac Goldstein"
date: "2023-06-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

```
# Libraries
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(tidybayes)
library(rstan)
library(posterior)
set.seed(236)
options(mc.cores = parallelly::availableCores())
rstan_options(auto_write = TRUE)
```

# Data
```{r}
rat_data = read.table("https://raw.githubusercontent.com/vnminin/sismid_mcmc_one/master/2022/code/rat_tumor.txt", header=TRUE)

num_obs = dim(rat_data)[1]
```

# Fitting the model using rstan
The following code fits the beta_binom.stan model to the rat tumor data.
The posterior_object contains both the data, as well as parameters specifying the model priors.
These are all treated as "data" by stan.
```{r message = FALSE, warning = FALSE}
posterior_object <- list(priors_only = FALSE,
                         num_obs = num_obs,
                         x = rat_data$x,
                         n = rat_data$n,
                         lambda_alpha = 0.1,
                         lambda_beta = 0.1)

chain = 4
iterations = 1000
seed = 236
thin = 1

posterior_fit <- stan(file = "src/beta_binom.stan",
                  data = posterior_object,
                  seed = seed,
                  iter = iterations,
                  thin = thin,
                  chain = chain)

```


# Assessing MCMC Diagnostics

We can assess the traceplot for the scaled likelihood.
```{r}
traceplot(posterior_fit, pars = "lp__")

```

We can also summarise Rhat and effective sample size.
```{r}
summary <- posterior_fit %>% 
           as_draws() %>% 
           summarise_draws() %>% 
           summarise(
             min_rhat = min(rhat),
             max_rhat = max(rhat),
             min_ess_bulk = min(ess_bulk),
             max_ess_bulk = max(ess_bulk),
             min_ess_tail = min(ess_tail),
             max_ess_tail = max(ess_tail)
           )

summary
```

# Visualizing the results
```{r fig.width = 10, fig.height = 6}
theta_draws <- posterior_fit %>% spread_draws(theta[i])

vis_theta <- theta_draws %>% filter(i %% 2 == 1) %>%
             arrange(i) %>%
             mutate(theta_lab = paste("Theta ", i)) 


theta_plot <- vis_theta %>% 
              ggplot(aes(x = fct_inorder(theta_lab), y = theta)) +
              geom_boxplot() +
              theme_bw() +
              xlab("Theta") +
              ylab("Value") + 
              theme(axis.text.x = element_text(angle = 90),
                    text = element_text(size = 20))

theta_plot
```

```{r}
fixed_vars <- posterior_fit %>% spread_draws(alpha, beta) 

plot_vars <- fixed_vars %>% pivot_longer(-c(.chain, .iteration, .draw), names_to = "name")

fixed_plot <- plot_vars %>% 
              ggplot(aes(x=value)) +
              geom_histogram() +
              facet_wrap(vars(name)) +
              theme_bw() +
              xlab("Value") + 
              theme(text = element_text(size = 20))

fixed_plot
```

